High geomagnetic field intensity recorded by anorthosite xenoliths requires a strongly powered late Mesoproterozoic geodynamo

### 2022-04-04 PNAS_manuscript/PNAS_Revision_Letter.tex

\documentclass[11pt, letterpaper]{article}
\usepackage{xcolor}
\usepackage{textcomp,marvosym}
\usepackage{amsmath,amssymb}
\usepackage[left]{lineno}
\usepackage{changepage}
\usepackage{rotating}
\usepackage{natbib}
\usepackage{setspace}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage{sidecap}
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\usepackage{titlesec}

@@

These are good points. We have added a sentence to the introduction that briefly describes that there are proposals for additional power sources to sustain the geomagnetic field including exsolution of Mg and Si-bearing minerals at the core-mantle boundary. We think that the discussion of plate tectonic driven modulation to field strength is better positioned within the discussion than the introduction. We have added additional detail regarding such hypotheses.

@@

We agree with the overall take-away that our new data are clearly at odds with the polynomial fit of Bono et al. (2019). While additional statistical treatments of compilations are available (e.g. Biggin et al., 2015), in the present manuscript we think it best to focus on these new paleointensity estimates and their relative strength to highlight this main result. With these questions being well-positioned for further thought down the line as the reviewer writes.

We have added additional details on the timescale of plate tectonic drive fluctuations in heat flows with connections to mantle dynamics timescales.

@@

The text ``due to the exclusion of light elements'' is added to be more specific about the mechanism behind compositional bouynacy.

@@

We have added ``far from the continental margin and subsequent orogenesis.''

@@

The paragraph following line 97 explains the uniqueness in terms of its distinct petrologic and magnetic properties from other anorthosite cumulus rocks. To avoid overstating these properties we changed this sentence to ``In this study, we target the high-purity anorthosite xenoliths of the Beaver River diabase in the Midcontinent Rift."

@@

In fact, plagioclase often bears small amount of Fe but weathering under surface conditions are unlikely to produce iron oxides that affect the magnetic minerals significantly. We have split the sentence into two: ``They are attractive targets for paleomagnetic study as plagioclase crystals can enshroud magnetic inclusions from alteration. In addition, the weathering of the plagioclase crystals do not resulting in the formation of secondary iron oxides, whereas other minerals such as olivine and pyroxene are more likely to alter and influence the magnetic mineralogy."

@@

We added ``from tholeiitic magma temperatures" to this sentence and added ``The details regarding the thermal history of these anorthosite xenoliths are discussed in \cite{Zhang2021b}" to help refer readers to our previous work.

@@

Text changed to ``22 {plus minus} 2{degree sign}."

@@

@@

@@

We decide to keep the Bz scale of the QDM maps as is. Bringing them to the same scale will either diminish the dipoles in the map of the Beaver River anorthosite xenolith or saturate the color scales in the Duluth anorthosite map. For clarification we noted in the figure caption that the two maps have Bz scales an order of magnitude different.

@@

Again thermal modeling suggests that the anorthosite and diabase cooled together from tholeiitic temperature. For this clarification reason, we would like to keep the short section of ``Background on anorthosite xenoliths of the Beaver River diabase" which summarizes previous geochronology, paleomagnetic direction, and thermal modeling results in \cite{Zhang2021b}.

@@

This is a good point. We did not make it clear in the text. 500\textdegree C is an estimated lab unblocking temperature for the anorthosites. Considering the cooling model of \cite{Zhang2021a} and the effect of lowering unblocking temperature due to slow cooling \cite{Halgedahl1980a}, a conservative estimate for the cooling time for the anorthosite from the Curie temperature of magnetite to the natural unblocking temperature should be about 10 kyr, leading to a cooling rate of 2.5\times10^{-10}. This is the value we used for calculating the paleointensity estimate value of 36\%. This was not made clear in the original manuscript.

@@

Thanks for doing this calculation. We have changed the text to ``From the thermal history model of ref.  \citealp{Zhang2021b}, we can estimate the duration over which the diabase and anorthosite cooled from the Curie temperature of magnetite ($\sim$580\textdegree C) to the time when they blocked in the majority of their characteristic natural remanence magnetization (lab unblocking temperatures $\sim$500\textdegree C;  \citealp{Zhang2021b}). We find the cooling time to be on the order of $\sim$10 kyr considering the effect of slow cooling on lowering the natural remanence unblocking temperatures \cite{Halgedahl1980a}. This range of cooling time corresponds to a cooling rate of $\sim2.5\times10^{-10}$ $^\circ$C s$^{-1}$. In contrast, the lab cooling rate is much faster through the same temperature interval with an estimated cooling rate of $\sim1.3\times10^{-1}$ $^\circ$C s$^{-1}$. The significant cooling rate difference leads to a predicted $\sim$36\% overestimate of true ancient field following the model of \citealp{Halgedahl1980a} (Fig. S3)."

@@

@@

@@

@@

@@

Thanks for pointing this out. This MAD selection criteria needed to be updated. The data in fact have much smaller scatter, it is just the selection criteria we implemented were more generous than necessary. A MAD selection of $<$10 would result in the same selection as the highest specimen MAD is around 8.

@@

@@

An additional supplementary figure S6 which plots the fitted coercivity curve and the smoothed data points has been added. The font sizes have been increased.

@@

Thanks for catching this inconsistency. Considering the effect of lowering unblocking temperature due to slow cooling, the cooling timescale should be closer to 10 kyr. We have changed the text to ``We find the cooling time to be on the order of $\sim$10 kyr considering the effect of slow cooling on lowering the natural remanence unblocking temperatures \cite{Halgedahl1980a}."

@@

We have added text ``This factor may result in an over-correction, as some multidomain grains that do not tend to severely bias paleointensity estimates toward higher \cite{Biggin2013a}, may exist in the anorthosite xenoliths as is evidenced through the slightly zigzagging Arai plots" to the related paragraph.

@@

Figure S5 which compares our samples and resampled Cenozoic paleointensity records illustrates that our values are also near the high end.

@@

>>>>

\documentclass[11pt, letterpaper]{article}
\usepackage{xcolor}
\usepackage{textcomp,marvosym}
\usepackage{amsmath,amssymb}
\usepackage[left]{lineno}
\usepackage{changepage}
\usepackage{rotating}
\usepackage{natbib}
\usepackage{setspace}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage{sidecap}
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\usepackage{titlesec}
\usepackage{url}

@@

These are good points. We have added a sentence to the introduction that briefly describes that there are proposals for additional power sources to sustain the geomagnetic field including exsolution of Mg and Si-bearing minerals at the core-mantle boundary. We think that the discussion of plate tectonic driven modulation to field strength is better positioned within the discussion than the introduction.

@@

We agree with the overall take-away that our new data are clearly at odds with the polynomial fit of \cite{Bono2019a}. While additional statistical treatments of compilations are available \cite[e.g.][]{Biggin2015a}, in the present manuscript we think it best to focus on these new paleointensity estimates and their relative strength to highlight this main result. These questions are well-positioned for further analysis down the line as the reviewer writes.

@@

The text ``due to the exclusion of light elements'' is added to be more specific about the mechanism behind compositional buoyancy.

@@

We have added ``far from the continental margin and subsequent orogenesis."

@@

The paragraph following line 97 explains the distinct petrologic and magnetic properties of the anorthosite xenoliths. To avoid overstating we changed this sentence to ``In this study, we target the high-purity anorthosite xenoliths of the Beaver River diabase in the Midcontinent Rift."

@@

We have split the sentence into two: ``They are attractive targets for paleomagnetic study as plagioclase crystals can enshroud magnetic inclusions from alteration. In addition, the weathering of the plagioclase crystals do not resulting in the formation of secondary iron oxides, whereas other minerals such as olivine and pyroxene are more likely to alter and influence the magnetic mineralogy."

@@

We have integrated this paragraph with the section of ``Background on anorthosite xenoliths of the Beaver River diabase" and the paragraph now reads ``In this study, we target the high-purity anorthosite xenoliths of the Beaver River diabase in the Midcontinent Rift. While magmatic activity within the Midcontinent Rift was protracted, there were intervals of particularly rapid volcanism and voluminous emplacement of intrusions \cite{Swanson-Hysell2021a}. The ca. 1092 Ma Beaver Bay Complex in northern Minnesota punctuates one such period of magmatism during the main stage of Midcontinent Rift activity. The magma that formed the 1091.7$\pm$0.2 Ma Beaver River diabase dikes and sills of the Beaver Bay Complex transported numerous anorthosite xenoliths that have short-axis diameters up to 180 meters via wide conduits \cite{Boerboom2004a, Boerboom2006b}. These anorthosite xenoliths are plagioclase cumulates that formed comagmatically with the host diabase in the lower crust---an interpretation confirmed by U-Pb zircon geochronology \cite{Zhang2021b}. Anorthosites are cumulate rocks composed dominantly of the mineral plagioclase. They are attractive targets for paleomagnetic study as plagioclase crystals can enshroud magnetic inclusions from alteration. In addition, the weathering of the plagioclase crystals do not resulting in the formation of secondary iron oxides, whereas other minerals such as olivine and pyroxene are more likely to alter and influence the magnetic mineralogy. The anorthosite xenoliths targeted in this study were brought to the near surface in magma that formed hypabyssal (shallowly emplaced) intrusions of the Beaver River diabase \cite{Zhang2021b}. They were likely heated to tholeiitic magma temperatures and partially recrystallized \cite{Zhang2021b}. Fe-Ti oxides within the plagioclase exsolved above magnetite Curie temperature ($\sim$580\textdegree C; \cite{Bian2021a}) and subsequently cooled and acquired thermal remanent magnetizations in conjunction with the host diabase at a paleolatitude of 22 $\pm$ 2\textdegree (calculated from the paleomagnetic pole of the coeval Portage Lake Volcanics; \citealp{Swanson-Hysell2019a}). The details regarding the thermal history of these anorthosite xenoliths are discussed in \cite{Zhang2021b}. Paleointensity experiments on the anorthosite xenoliths have a high success rate, yielding consistent specimen- and site-level paleointensity results. The anorthosite xenoliths show low anisotropy of thermal remanent magnetization (TRM) acquisition and can acquire TRM linearly within the range of relevant field strengths. Magnetic imaging shows that the anorthosite specimens have dominant magnetic carriers within and interstitial to plagioclase crystals without strong preferred orientations. Thermal modeling results and paleomagnetic directional data show the anorthosite xenoliths to have acquired thermal remanent magnetizations while cooling with the Beaver River diabase \cite{Zhang2021b}. Step-wise thermal demagnetization data show the anorthosite xenoliths to have dominantly single-component magnetization that often unblocks sharply within temperature ranges between 500\textdegree C and 580\textdegree C, consistent with magnetization held by low-titanium titanomagnetite \cite{Zhang2021b}. "

@@

Text changed to ``22 $\pm$ 2\textdegree."

@@

We removed this sentence.

@@

We have integrated this section into the introduction.

@@

We decide to keep the Bz scale of the QDM maps as is. Bringing them to the same scale will either diminish the dipoles in the map of the Beaver River anorthosite xenolith or saturate the color scales in the Duluth anorthosite map. For clarification, we noted in the figure caption that the two maps have Bz scales that are an order of magnitude different.

@@

We clarify this in the introduction.

@@

This is a good point. We did not make it clear in the text. 500\textdegree C is an estimated lab unblocking temperature for the anorthosites based on thermal demagnetization of \cite{Zhang2021b} and the zero-field heating steps of this study. Considering the cooling model of \cite{Zhang2021a} and the effect of lowering unblocking temperature due to slow cooling \cite{Pullaiah1975a}, an estimate for the cooling time for the anorthosite from the Curie temperature of magnetite to the natural unblocking temperature could be longer than 1 kyr. To fully account for the range of various lower temperature bounds used for optimize the paleointensity estimate, we use 10 kyr for calculating the cooling rate correction factor. This leads to a cooling rate of 2.5\times10$^{-10}$. This is the value we use for calculating the paleointensity estimate value of 36\%.

@@

Thanks for doing this calculation. We have changed the text to ``To fully account for the range of various lower temperature bounds used for optimize the paleointensity estimate and the effect of lowering unblocking temperature due to slow cooling, we estimate the anorthosite cooled over a time period on the order of $\sim$10 kyr (from $\sim$580\textdegree C to as low as $\sim$400\textdegree C) based the thermal history model of ref. \citealp{Zhang2021b} and cooling rate correction model of \cite{Pullaiah1975a}. This cooling time corresponds to a cooling rate of $\sim2.5\times10^{-10}$ $^\circ$C s$^{-1}$. In contrast, the lab cooling rate is much faster through the same temperature interval with an estimated cooling rate of $\sim1.3\times10^{-1}$ $^\circ$C s$^{-1}$."

@@

While the independent two-sample t-test can provide statistical assessment for the similarity of site- and region-level results, the difference between high values from the Kennedy Creek region and the other two regions are quite significant, and the overlapping uncertainties between site AX12 and AX16, and AX11 and AX13 are indicative of their distribution similarities.

We have added reference to our table S3 and added statement of t-test results within and between regions in the caption for table S3. Independent two-sample t-test results show that site AX12 and AX16 pass the t-test at 1\% significance level; site AX11 and AX13 pass the test at 5\% significance level; region Kennedy Creek have distinguishable distributions from the other two regions; Silver Bay region and Carlton Peak region are indistinguishable at 5\% significance level.

@@

We agree with the comments and have added additional text to the introduction section regarding the light element precipitation aspect.

@@

To aid the reader to identify the timing of these transitions, we have added more ticks on the x axis of figure 6. This should make the figure and related text easier to follow.

@@

We think that the discussion of plate tectonic driven modulation to field strength is better positioned within the discussion than the introduction.

@@

Thanks for pointing this out. This MAD selection criteria needed to be updated. The data in fact have much smaller scatter, it is just the selection criteria we implemented were more generous than necessary. A MAD selection of $<$10 would result in the same selection as the highest specimen MAD is around 8. Our table Figure S3 also presents the small scatter of MAD.

@@

While additional inset figures can help illustrating the dipole directions they would add too much to the magnetic field maps and complicate the figure. In addition, it would make more sense to show a population of magnetic sources all rotate as the external field changes direction Our figure caption should help guide readers to understand these maps better. We hope our caption  ``The magnetic images show that remanent magnetizations of Beaver River anorthosite (i.e. the individual dipoles visible with paired red +B$_z$ and blue -B$_z$ lobes) align well with the first applied field direction (E) and then rotated to align with the second applied field direction indicating minimal anisotropic behavior (G)" describes the rotation of the dipoles in detail.

@@

The three example specimen shown in this figure actually have quite nice data that are not very noisy. To make the main figure clear and highlight the difference in peak coercivities, we decide to keep the presentation of this figure as is. The font sizes have been increased.

An additional supplementary figure S6 which plots the fitted coercivity curve and the smoothed data points has been added.

@@

Thanks for catching this inconsistency. Considering the effect of lowering unblocking temperature due to slow cooling, the cooling timescale should be closer to 10 kyr. We have changed related text to ``To fully account for the range of various lower temperature bounds used for optimize the paleointensity estimate and the effect of lowering unblocking temperature due to slow cooling, we estimate the anorthosite cooled over a time period on the order of $\sim$10 kyr (from $\sim$580\textdegree C to as low as $\sim$400\textdegree C) based the thermal history model of ref. \citealp{Zhang2021b} and cooling rate correction model of \cite{Pullaiah1975a}. This cooling time corresponds to a cooling rate of $\sim2.5\times10^{-10}$ $^\circ$C s$^{-1}$. In contrast, the lab cooling rate is much faster through the same temperature interval with an estimated cooling rate of $\sim1.3\times10^{-1}$ $^\circ$C s$^{-1}$. The significant cooling rate difference leads to a predicted $\sim$36\% overestimate of true ancient field following the model of \citealp{Pullaiah1975a} (Fig. S3)."

@@

We have added text ``This factor may result in an over-correction, as some multidomain and pseudo-single domain grains that do not tend to severely bias paleointensity estimates toward higher \cite{Biggin2013a}, may exist in the anorthosite xenoliths as is evidenced through the slightly zigzagging Arai plots" to the related paragraph.

@@

Figure S5 which compares our samples and resampled Cenozoic paleointensity records illustrates that our values are also near the high end. We discuss that ``we perform the same comparison with the existing paleointensity data for the Cenozoic Era (the last 66 million years) from the PINT database (filtered by Q$_{PI}\geq$3; Fig. S5) given that the observational dataset could record more geomagnetic field excursions or regional high flux patches. The anorthosite xenoliths 3 region-mean intensity is higher than all values from the resample results of the PADM2M model, and rival the top 3\% values of the resample averages from the Cenozoic data (Fig. S5)."

@@

\bibliographystyle{gsabull}
\bibliography{YZ_ref}
